---
title: "Analysis of the Value-Weight Ratio of Commercial Landings in North Carolina
  Counties, 2014 - 2022"
author: "Ayden Schirmacher, Natalie Holsclaw, Tori Newton, and Julia Kagiliery"
output:
  html_document:
    df_print: paged
---
\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("lubridate")
#install.packages("ggrepel")
#install.packages("here")
#install.packages("gifski")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("scales")
#install.packages("sf")
#install.packages("leaflet")
#install.packages("mapview")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggrepel)
library(here)
library(gifski)
library(gganimate)
library(gapminder)
library(scales)
library(sf)
library(leaflet)
library(mapview)

# Set your ggplot theme

# Load your datasets

```

```{r data import, message=FALSE, warning=FALSE, include=FALSE}
getwd()

#catch and species information
fish_2022<-read.csv(here("Data", "Raw", "2022.csv"))
fish_2021<-read.csv(here("Data", "Raw", "2021.csv"))
fish_2020<-read.csv(here("Data", "Raw", "2020.csv"))
fish_2019<-read.csv(here("Data", "Raw", "2019.csv"))
fish_2018<-read.csv(here("Data", "Raw", "2018.csv"))
fish_2017<-read.csv(here("Data", "Raw", "2017.csv"))
fish_2016<-read.csv(here("Data", "Raw", "2016.csv"))
fish_2015<-read.csv(here("Data", "Raw", "2015.csv"))
fish_2014<-read.csv(here("Data", "Raw", "2014.csv"))

fish_all <- bind_rows(
  "2014" = fish_2014, 
  "2015" = fish_2015, 
  "2016" = fish_2016, 
  "2017" = fish_2017, 
  "2018" = fish_2018, 
  "2019" = fish_2019, 
  "2020" = fish_2020, 
  "2021" = fish_2021, 
  "2022" = fish_2022, 
  .id = "Year")


#spatial and county data
fish_10year<-read.csv(here("Data", "Raw", "stats_fishies.csv"))
fish_10year

```

```{r SPECIES data cleaning, message=FALSE, warning=FALSE, include=FALSE}

fish_all$FINALWEIGHT <- gsub("[,]", "", fish_all$FINALWEIGHT) 
#^^remove commas from data
fish_all$FINALVALUE <- gsub("[,]", "", fish_all$FINALVALUE)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("\\$", "", fish_all$FINALVALUE)
#^^remove dollar signs from data

fish_all<-fish_all%>%
  mutate(FINALWEIGHT=as.integer(FINALWEIGHT))%>% #make data into numerical
  mutate(FINALVALUE=as.integer(FINALVALUE))%>% #make data into numerical
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  #^^create date object
  filter(!is.na(FINALWEIGHT))%>% #filter for NA
  filter(!is.na(FINALVALUE)) #filter for NA

#Keeping only the fish observations that are present across all years
species_clean<-fish_all %>%
  select(Year, FISH, FINALWEIGHT, FINALVALUE)%>%
  group_by(FISH) %>%
  filter(n_distinct(Year) == length(unique(fish_all$Year))) %>%
  #^^keep only the observations across all years
  ungroup()
```

```{r COUNTY data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#group data by year and county
county_clean<-fish_10year%>%
  filter(!is.na(Pounds))%>% #filter for NA
  filter(!is.na(Value))%>% #filter for NA
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  group_by(County)%>%
  filter(n_distinct(Year) == length(unique(fish_10year$Year)))
  #^^keep only observations in all years
```

```{r SPECIES data group by species type, message=FALSE, warning=FALSE, include=FALSE}

species_clean <- species_clean %>%
  mutate(species_group = case_when(
    FISH %in% c("Blue Crabs, Hard", "Blue Crabs, Peeler", 
                "Blue Crabs, Soft", "Stone Crabs", ("Shrimp (Heads On)")) 
    ~ "Crustaceans",
    
    FISH %in% c("Clams, Hard (Meats)", "Clams, Hard (Numbers)", 
                "Oysters (Bushels)", "Oysters (Meats)", 
                "Scallop, Sea (Meats)", "Squid", 
                "Whelks/Conchs (Meats)", "Octopus") 
    ~ "Mollusks",

    FISH %in% c("Dogfish, Smooth", "Dogfish, Spiny", 
                "Sharks", "Skates") 
    ~ "Sharks & Rays",

    FISH %in% c("Eel, American") ~ "Eels",

    FISH %in% c("Flounder, Southern", "Flounder, Summer", 
                "Flounders, Other") 
    ~ "Flatfish",

    FISH %in% c("Croaker, Atlantic", "Drum, Black", "Drum, Red", 
                "Seatrout, Spotted", "Weakfish (Grey Trout)", 
                "Spot") 
    ~ "Drum & Croaker Family",

    FISH %in% c("Grouper, Gag", "Grouper, Red", "Grouper, Scamp", 
                "Grouper, Snowy", "Groupers, Other", "Sea Basses") 
    ~ "Groupers & Sea Basses",

    FISH %in% c("Snapper, Vermilion (Beeliner)", 
                "Snappers, Other", "Porgies") 
    ~ "Snappers & Porgies",

    FISH %in% c("Mackerel, Atlantic (Boston)", "Mackerel, King", 
                "Mackerel, Spanish", "Tuna, Bigeye", "Tuna, Bluefin",
                "Tuna, Yellowfin", "Tunas, Other", 
                "Tunny, Little (False Albacore)") ~ "Mackerels & Tunas",

    FISH %in% c("Amberjacks", 
                "Jacks (Crevalle, Rainbowrunner, Bluerunner)", "Pompano") 
    ~ "Jacks & Pompanos",

    FISH %in% c("Anglerfish (Monkfish Including Monklivers)", "Bluefish", "Bonito", 
                "Butterfish", "Carp", "Catfishes", "Cobia", "Cutlassfish, Atlantic", 
                "Dolphinfish", "Garfish", "Grunts", "Hakes", "Harvestfish (Starbutters)", 
                "Hogfish (Hog Snapper)", "Mullet, Sea (Kingfishes)", "Mullet, Striped", 
                "Perch, White", "Perch, Yellow", "Pigfish", "Pinfish", "Pufferfish", 
                "Scup", "Shad, American", "Shad, Gizzard", "Shad, Hickory", "Sheepshead", 
                "Skippers", "Spadefish", "Striped Bass", "Swordfish", "Tilefish, Blueline", 
                "Tilefish, Other", "Triggerfish", "Wahoo") ~ "Other Finfish",
    FISH %in% c("Unclassified Fish For Industrial/Bait", "Unclassified Fish for Food", 
                "Unclassified Shellfish") ~ "Unclassified",
    TRUE ~ "Other"
  ))

```

```{r SPECIES data new value to weight ratio, message=FALSE, warning=FALSE, include=FALSE}
species_clean<-species_clean%>%
  mutate(value_weight_ratio=FINALVALUE/FINALWEIGHT)
```

```{r SPECIES data create difference in means between 1972 and 2022, message=FALSE, warning=FALSE, include=FALSE}
species_year01<-species_clean%>%
  filter(Year=="2014-01-01")
species_year10<-species_clean%>%
  filter(Year=="2022-01-01")

species_diff<- 
  left_join(species_year10, species_year01, by = "FISH", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

species_clean <- left_join(species_clean, species_diff %>% 
                        select(FISH, diff_means), by = "FISH")

```

```{r COUNTY data create difference in means between 2014 and 2022, message=FALSE, warning=FALSE, include=FALSE}
county_clean<-county_clean%>%
  mutate(value_weight_ratio=Value/Pounds)

county_year01<-county_clean%>%
  filter(Year=="2014-01-01")
county_year10<-county_clean%>%
  filter(Year=="2022-01-01")

county_diff<- 
  left_join(county_year10, county_year01, by = "County", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

county_clean <- left_join(county_clean, county_diff 
                      %>% select(County, diff_means), by = "County")

```


# Rationale and Research Questions

## Question 1 : Is there a significant difference in the weight-value ratio from commercial landings across species groups, time, and counties in North Carolina between 2014 and 2022?



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 

*Analysis By Species Group*

  An analysis of the average value-to-weight ratio by species over the study period shows that some of the species groups have little to no variation, while others are much more scattered (Figure 1.) For example, the value-to-weight ratio for Crustaceans experiences a change of~5 between 2021 and 2022 and Eels experienced a change of ~3.5 between 2017 and 2019. However, the ratio for Sharks and Rays has stayed constant just above zero, while many other species groups have had small but steady positive increases (small positive slopes).
  
  Another way of looking at the value-to-weight ratio is by plotting the ratio itself, without distinguishing the year of observation. The relationship of value-to-weight is shown in Figure 2. Due to the large right skew from the Crustaceans species group in this data, the actual relationships of the species groups are difficult to see. Therefore, a log transformation was applied in order to better distinguish the value-to-weight ratios of the different species groups (Figure 3.). From these two plots, it can be determined that Crustaceans tend to be an outlying species group; the lowest value-to-weight ratios typically correspond to mollusks; and that many of the other species have fairly similar ratios.
  
  A final analysis of the value-to-weight ratio by species group is shown in a year-by-year plot of the weight-to-value ratio. This shows the same information as Figures 2. and 3., but now differentiates by year. The x axis (weight) is log transformed in order to better visualize the differences between species groups -- otherwise, most of the data is too small compared to crustaceans to be useful. Outlying values are labeled for every year. This plot confirms the outlying crustaceans, but also shows which years have certain species groups experiencing a ratio increase. An increasing ratio indicates that the value of the species landings has increased faster than the weight of the landings (where value-to-weight ratio is calculated by Value/Weight) (Figure 4.).
  
  Overall, the exploratory plots for species groupings indicate that there is a high likelihood of the value-to-weight ratio of commercial landings in North Carolina being different between 2014 and 2022, and across species groups throughout that time period. 
  
```{r mean ratio across species groups, warning=FALSE, message=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Figure 1. Average Weight-Value Ratio of Commercial Landings in North Carolina by Species Group, 2014-2022'}

# Mean ratio ($) across species groups 
summary_table <- species_clean %>%
  group_by(Year, species_group) %>%
  summarize(Mean = mean(value_weight_ratio, na.rm = TRUE),
            Median = median(value_weight_ratio, na.rm = TRUE),
            SD = sd(value_weight_ratio, na.rm = TRUE))

# Print the summary table
print(summary_table)

# Visualization 
ggplot(summary_table, aes(x = Year, y = Mean, color = species_group)) +
  geom_line() +  # Line plot for the mean ratio over time
  geom_point() +  # Add points to show individual data points
  labs(
    title = "Mean Weight-Value Ratio by Species Group over Time",
    x = "Time",
    y = "Mean Weight-Value Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 
 #+ facet_wrap(~species_group)
  ####!!! without the facet wrap the graph seems too busy, but up to you to decide!!
```




```{r SPECIES data ratio plot, warning=FALSE, message=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Figure 2. Value to Weight Ratio of Commercial Landings in North Carolina by Species Group, 2014-2022'}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value by Species Type",
    x="Weight (lbs)", y="Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```




```{r SPECIES data log ratio plot, warning=FALSE, message=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Figure 3. Log Transformed Value to Weight Ratio of Commercial Landings in North Carolina by Species Group, 2014-2022'}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  scale_x_log10(labels=comma)+
  scale_y_log10(labels=comma)+
  geom_point(alpha = 0.40)+
  ggtitle("Logged Ratio of Fish Weight to Value by Species Type")+
  xlab("Log Scale Weight (lbs)")+
  ylab("Log Scale Value ($USD)")+
  theme_bw()

```




```{r SPECIES data animation over time, warning=FALSE, message=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Figure 4. Annual Value to Log Weight Ratio of Commercial Landings in North Carolina by Species Group, 2014-2022, where annual outliers are labeled'}
species_animate<-
  ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, 
         size=3,
         col = FISH)) +
    geom_point(alpha = 0.7, show.legend = FALSE) +
    scale_x_log10(labels=comma)+
    scale_y_continuous(labels=comma)+
    labs(title = 'Ratio of Value to Log Weight, Year: {closest_state}', 
        x = 'Log Scale Weight (Units)', 
        y = 'Value ($USD)') +
    transition_states(Year, transition_length = 1, state_length = 1,
                      wrap = TRUE)+
    ease_aes('cubic-in-out')+
    geom_text(data = species_clean%>%
      filter(FINALVALUE > 5000000),
      aes(label=FISH), 
      hjust = 1, vjust = 0, size = 3, color = "black")

animate(species_animate, fps = 30, duration = 15)
```


*Analysis By County*
  The commercial landings data can also be classified by county. Figure 5 shows the value-to-weight ratio by county across all years in the study period; from this plot it can be determined that not only is there a difference between counties, but within each county, there was an overall increase in the value-to-weight ratio during the study period. Again, this indicates that there is an larger increase in the value of the landings than there is an increase in the weight of the landings. 

```{r COUNTY shapefile and joining dataset, warning=FALSE, message=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Figure 5. Value to Weight Ratio of Commercial Landings in North Carolina by County, 2014-2022'}
nc_county<-st_read(here("./Data/Processed/Spatial/cb_2018_us_county_20m.shp"))%>%
  filter(STATEFP == "37")

county_clean<-county_clean%>%
  mutate(NAME=County)

nc_county<-nc_county%>%
  inner_join(county_clean, by="NAME")

ggplot(nc_county)+
  geom_sf(aes(fill = value_weight_ratio))+
    labs(
    title= "Weight-Value Ratio by County of Landing")+
  theme_bw()+
  facet_wrap(~Year)
```

\newpage

## Analysis

# Main Research Question: Is there a difference in the value-to-weight ratio from commercial landings in North Carolina between 2014-2022?

# Sub-question A: Is there a difference in the value-to-weight ratio from commercial landings by SPECIES in North Carolina between 2014-2022?

# Sub-question B: Is there a difference in the value-to-weight ratio from commercial landings by COUNTY in North Carolina between 2014-2022?

#Null hypothesis for all questions: there is no difference in the value-to-weight ratio. 


*Addressing sub-question A*

  Using a multiple linear regression model, it can be determined that there are some significant differences in the value-to-weight ratios across species, holding all else constant and assuming that the impacts of the explanatory variable on the annual value-to-weight ratio apply for the study time period of 2014-2022. The value-to-weight ratio of this model is log transformed in order to better satisfy the normality assumptions of mulitple linear regression. 
  
  The regression model uses the Crustacean species group as the baseline, therefore we interpret all estimates (coefficients) and p-values in relationship to the Crustacean intercept. The species which are significantly different from the value-to-weight ratio of Crustaceans are: Drum and Croaker; Groupers and Sea Basses, Jacks and Pompanos, Mackerels and Tunas, Other, Other Finfish, Sharks and Rays, and Unclassified. All of these species groups have p-values less than 0.008, though only Grouper and Sea Basses have a positive relationship: an increase in the value-to-weight ratio of Crustaceans indicates and exp(0.485)≈1.625 or 62.5% increase in the Grouper and Sea Basses species group.
  
  The year is also a significant contributor to the value-to-weight ratio, with a p-value of ~0.007.
  
  Given the R-squared value, the model explains ~33.44% of the variations in value-to-weight ratios. While this may seem low, the model is also statistically significant with a p-value less than 0.001. The assumptions of the model are not perfect and likely have issues with heteroscedasticity based on the Residuals plot. The residuals v. fitted plot shows values clumped together but relatively evenly distributed above and below the line. 
  
  Finally, the F-statistic of the model appears to be relatively small at 27.63 on 13 and 715 degrees of freedom, indicating that the differences in value-to-weight ratio between groups are not entirely by chance. Alternatively, the F-statistics of the ANOVA test is 29.320. Given that the multiple regression ANOVA F-value represents the the ratio of the between-group variance to the within-group variance, having an F-value of 29.320 against 12 degrees of freedom on a p-value <0.001 suggests strong evidence to reject the null hypothesis that the value-to-weight ratio across species groups are the same. ******Interpret TUKEYs TEST, DISCUSS OUTLIERS??*********
  
  It can be concluded that there are significant differences in the value-to-weight ratio across species, holding all else constant. Therefore, we can reject the null hypothesis that there is no difference in the value-to-weight ratio between 2014-2022 across species. However, some issues may have disrupted these results. Issues include: potential heteroscedasticity of the data; there are also likely issues with spatial and temporal autocorrelation of the data. 

```{r SPECIES data multiple linear regression model, message=FALSE, warning=FALSE, include=FALSE}
species_multi_lm<-lm(data = species_clean, log(value_weight_ratio) ~ species_group + Year)

fish_anova <- aov(log(value_weight_ratio) ~  species_group + Year, data = species_clean)
summary(fish_anova)
# There is a significant different in value across species groups.
#fish_tukey <- TukeyHSD(fish_anova)
#fish_tukey
```

```{r SPECIES data multiple linear regression model output, message=FALSE, warning=FALSE, echo=FALSE, fig.align='left', fig.height=4, fig.cap='Table 1. Multiple Linear Regression of Logged Value to Weight Ratio of Commercial Landings in North Carolina by Species, 2014-2022'}
summary(species_multi_lm)
plot(species_multi_lm)
summary(species_multi_lm)
```


*Addressing Sub-question B*
***NOT SURE IF WE WANT TO HAVE A STATISTICAL DIFFERENCE TEST FOR COUNTY DATA*****

\newpage

# Summary and Conclusions






\newpage

# References
<add references here if relevant, otherwise delete this section> 
