---
title: "SchirmacherAyden_FinalProject"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
##Data Wrangling
```{r clear environment, include=FALSE, warning=FALSE, message=FALSE}
remove(list=ls())
```

```{r packages, warning=FALSE, message=FALSE, include=FALSE}
#load packages
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggrepel)
library(here)
library(gifski)
library(gganimate)
library(gapminder)
library(scales)
library(sf)
library(leaflet)
library(mapview)
library(MASS)
library(lmtest)
library(mgcv)
```

```{r data import, message=FALSE, warning=FALSE, include=FALSE}
#load in all the data

getwd()

#catch and species information
fish_2022 <- read.csv(here("Data", "Raw", "2022.csv"))
fish_2021 <- read.csv(here("Data", "Raw", "2021.csv"))
fish_2020 <- read.csv(here("Data", "Raw", "2020.csv"))
fish_2019 <- read.csv(here("Data", "Raw", "2019.csv"))
fish_2018 <- read.csv(here("Data", "Raw", "2018.csv"))
fish_2017 <- read.csv(here("Data", "Raw", "2017.csv"))
fish_2016 <- read.csv(here("Data", "Raw", "2016.csv"))
fish_2015 <- read.csv(here("Data", "Raw", "2015.csv"))
fish_2014 <- read.csv(here("Data", "Raw", "2014.csv"))

fish_all <- bind_rows(
  "2014" = fish_2014,
  "2015" = fish_2015,
  "2016" = fish_2016,
  "2017" = fish_2017,
  "2018" = fish_2018,
  "2019" = fish_2019,
  "2020" = fish_2020,
  "2021" = fish_2021,
  "2022" = fish_2022,
  .id = "Year"
)

#spatial and county data
fish_10year <- read.csv(here("Data", "Raw", "stats_fishies.csv"))
```

```{r SPECIES data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#clean data

fish_all$FINALWEIGHT <- gsub("[,]", "", fish_all$FINALWEIGHT)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("[,]", "", fish_all$FINALVALUE)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("\\$", "", fish_all$FINALVALUE)
#^^remove dollar signs from data

fish_all <- fish_all %>%
  mutate(FINALWEIGHT = as.integer(FINALWEIGHT)) %>% #make data into numerical
  mutate(FINALVALUE = as.integer(FINALVALUE)) %>% #make data into numerical
  mutate(Year = as.Date(paste0(Year, "-01-01"), format = "%Y-%m-%d")) %>%
  #^^create date object, keep 01-01 for plotting
  filter(!is.na(FINALWEIGHT)) %>% #filter for NA
  filter(!is.na(FINALVALUE)) #filter for NA

#Keeping only the fish observations that are present across all years
species_clean <- fish_all %>%
  dplyr::select(Year, FISH, FINALWEIGHT, FINALVALUE) %>%
  group_by(FISH) %>%
  filter(n_distinct(Year) == length(unique(fish_all$Year))) %>%
  #^^keep only the observations across all years
  ungroup()
```

```{r COUNTY data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#group data by year and county
county_clean<-fish_10year%>%
  filter(!is.na(Pounds))%>% #filter for NA
  filter(!is.na(Value))%>% #filter for NA
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  group_by(County)%>%
  filter(n_distinct(Year) == length(unique(fish_10year$Year)))
  #^^keep only observations in all years
```


```{r SPECIES data group by species type, message=FALSE, warning=FALSE, include=FALSE}
#now get data by species classification
species_clean <- species_clean %>%
  mutate(species_group = case_when(
    FISH %in% c("Blue Crabs, Hard", "Blue Crabs, Peeler", 
                "Blue Crabs, Soft", "Stone Crabs", ("Shrimp (Heads On)")) 
    ~ "Crustaceans",
    
    FISH %in% c("Clams, Hard (Meats)", "Clams, Hard (Numbers)", 
                "Oysters (Bushels)", "Oysters (Meats)", 
                "Scallop, Sea (Meats)", "Squid", 
                "Whelks/Conchs (Meats)", "Octopus") 
    ~ "Mollusks",

    FISH %in% c("Dogfish, Smooth", "Dogfish, Spiny", 
                "Sharks", "Skates") 
    ~ "Sharks & Rays",

    FISH %in% c("Eel, American") ~ "Eels",

    FISH %in% c("Flounder, Southern", "Flounder, Summer", 
                "Flounders, Other") 
    ~ "Flatfish",

    FISH %in% c("Croaker, Atlantic", "Drum, Black", "Drum, Red", 
                "Seatrout, Spotted", "Weakfish (Grey Trout)", 
                "Spot") 
    ~ "Drum & Croaker Family",

    FISH %in% c("Grouper, Gag", "Grouper, Red", "Grouper, Scamp", 
                "Grouper, Snowy", "Groupers, Other", "Sea Basses") 
    ~ "Groupers & Sea Basses",

    FISH %in% c("Snapper, Vermilion (Beeliner)", 
                "Snappers, Other", "Porgies") 
    ~ "Snappers & Porgies",

    FISH %in% c("Mackerel, Atlantic (Boston)", "Mackerel, King", 
                "Mackerel, Spanish", "Tuna, Bigeye", "Tuna, Bluefin",
                "Tuna, Yellowfin", "Tunas, Other", 
                "Tunny, Little (False Albacore)") ~ "Mackerels & Tunas",

    FISH %in% c("Amberjacks", 
                "Jacks (Crevalle, Rainbowrunner, Bluerunner)", "Pompano") 
    ~ "Jacks & Pompanos",

    FISH %in% c("Anglerfish (Monkfish Including Monklivers)", "Bluefish", "Bonito", 
                "Butterfish", "Carp", "Catfishes", "Cobia", "Cutlassfish, Atlantic", 
                "Dolphinfish", "Garfish", "Grunts", "Hakes", "Harvestfish (Starbutters)", 
                "Hogfish (Hog Snapper)", "Mullet, Sea (Kingfishes)", "Mullet, Striped", 
                "Perch, White", "Perch, Yellow", "Pigfish", "Pinfish", "Pufferfish", 
                "Scup", "Shad, American", "Shad, Gizzard", "Shad, Hickory", "Sheepshead", 
                "Skippers", "Spadefish", "Striped Bass", "Swordfish", "Tilefish, Blueline", 
                "Tilefish, Other", "Triggerfish", "Wahoo") ~ "Other Finfish",
    FISH %in% c("Unclassified Fish For Industrial/Bait", "Unclassified Fish for Food", 
                "Unclassified Shellfish") ~ "Unclassified",
    TRUE ~ "Other"
  ))

```

```{r SPECIES data new value to weight ratio, message=FALSE, warning=FALSE, include=FALSE}
#value to weight ratio

species_clean <-species_clean %>%
  mutate(value_weight_ratio=FINALVALUE/FINALWEIGHT)
```


```{r SPECIES data create difference in means between 1972 and 2022, message=FALSE, warning=FALSE, include=FALSE}
species_year01<-species_clean%>%
  filter(Year=="2014-01-01")
species_year10<-species_clean%>%
  filter(Year=="2022-01-01")

species_diff<- 
  left_join(species_year10, species_year01, by = "FISH", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

species_clean <- left_join(species_clean, species_diff %>% 
                        dplyr::select(FISH, diff_means), by = "FISH")

```

```{r COUNTY data create difference in means between 2014 and 2022, message=FALSE, warning=FALSE, include=FALSE}
county_clean <- county_clean %>%
  mutate(value_weight_ratio = Value / Pounds)

county_year01 <- county_clean %>%
  filter(Year == "2014-01-01")
county_year10 <- county_clean %>%
  filter(Year == "2022-01-01")

county_diff <-
  left_join(county_year10,
            county_year01,
            by = "County",
            suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

county_clean <- left_join(county_clean,
                          county_diff %>%
                            dplyr::select(County, diff_means),
                          by = "County")
```

##Exploratory Plots -- Catch and Species Data

*FIGURE 1. Average Weight-Value Ratio by Species Over Time*
```{r mean weight-value ratio across species groups}
# Mean ratio ($) across species groups 
summary_table <- species_clean %>%
  group_by(Year, species_group) %>%
  summarize(Mean = mean(value_weight_ratio, na.rm = TRUE),
            Median = median(value_weight_ratio, na.rm = TRUE),
            SD = sd(value_weight_ratio, na.rm = TRUE))

# Print the summary table
print(summary_table)

# Visualization 
ggplot(summary_table, aes(x = Year, y = Mean, color = species_group)) +
  geom_line() +  # Line plot for the mean ratio over time
  geom_point() +  # Add points to show individual data points
  labs(
    title = "Mean Weight-Value Ratio by Species Group Over Time",
    x = "Time",
    y = "Mean Weight-Value Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 
```

*FIGURE 2. Catch Weight-to-Value Data Categorized by Species*
```{r SPECIES data ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value from Commercial Fishing
    by Species Type in North Carolina, 2014-2022",
    x="Weight (lbs)", y="Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```

*FIGURE 3.Catch Weight-to-Value Data Categorized by Species with Log Transformations on X and Y Axis*
```{r SPECIES data log ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  scale_x_log10(labels=comma)+
  scale_y_log10(labels=comma)+
  geom_point(alpha = 0.40)+
  ggtitle("Logged Ratio of Fish Weight to Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  xlab("Log Scale Weight (lbs)")+
  ylab("Log Scale Value ($USD)")+
  theme_bw()

#so this looks better but We will check for normality later 
```

*FIGURE 4. ANIMATED Annual Catch Weight-to-Value Data Categorized by Species with Log Transformations on X and Y Axis and Labels on Outlying Values*
```{r SPECIES data animation over time, warning=FALSE, message=FALSE, echo=FALSE}
species_animate<-
  ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, 
         size=3,
         col = FISH)) +
    geom_point(alpha = 0.7, show.legend = FALSE) +
    scale_x_log10(labels=comma)+
    scale_y_continuous(labels=comma)+
    labs(title = 'Ratio of Value to Log Weight, Year: {closest_state}', 
        x = 'Log Scale Weight (Units)', 
        y = 'Value ($USD)') +
    transition_states(Year, transition_length = 1, state_length = 1,
                      wrap = TRUE)+
    ease_aes('cubic-in-out')+
    geom_text(data = species_clean%>%
      filter(FINALVALUE > 5000000),
      aes(label=FISH), 
      hjust = 1, vjust = 0, size = 3, color = "black")

animate(species_animate, fps = 30, duration = 15)
```

*FIGURE 5. Catch Weight-to-Value Data Categorized by County*
```{r COUNTY shapefile and joining dataset, warning=FALSE, message=FALSE, include=FALSE}
nc_county<-st_read(here("./Data/Processed/Spatial/cb_2018_us_county_20m.shp"))%>%
  filter(STATEFP == "37")

county_clean<-county_clean%>%
  mutate(NAME=County)

nc_county<-nc_county%>%
  inner_join(county_clean, by="NAME")

ggplot(nc_county)+
  geom_sf(aes(fill = value_weight_ratio))+
    labs(
    title= "Weight-Value Ratio of Commercial Seafood Landings by
    County of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year) + 
  theme(
    axis.text.x = element_text(angle = 65, hjust = 1)
  )
```

*FIGURE 6. Catch Data County Spatial Plots Categorized by Value*
```{r COUNTY data spatial value plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = Value))+
    labs(
    title= "Value of Commercial Seafood Landings by County
    of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year) + 
  theme(
    axis.text.x = element_text(angle = 65, hjust = 1)
  )
```


##Statistical Analysis -- Catch and Species Data

```{r}
#here are summary stats by fish 
SummaryStatsFish <- fish_all |>
  mutate(ValueWeightRatio = FINALVALUE / FINALWEIGHT) |>  # First calculate the ratio for each row
  group_by(FISH) |>
  summarize(
    AverageWeight = mean(FINALWEIGHT, na.rm = TRUE),
    SDWeight = sd(FINALWEIGHT, na.rm = TRUE),
    AverageValue = mean(FINALVALUE, na.rm = TRUE),
    SDValue = sd(FINALVALUE, na.rm = TRUE),
    AverageValueWeightRatio = mean(ValueWeightRatio, na.rm = TRUE), 
    SDRatio = sd(ValueWeightRatio, na.rm = TRUE),# Calculate the mean of the ratio
    .groups = "drop"  # This removes the grouping structure after summarization
  )

print(SummaryStatsFish)
```

```{r}
SummaryStatsSpecies <- species_clean |>
  group_by(species_group) |>
  summarize(
    AverageWeight = mean(FINALWEIGHT, na.rm = TRUE),
    SDWeight = sd(FINALWEIGHT, na.rm = TRUE),
    AverageValue = mean(FINALVALUE, na.rm = TRUE),
    SDValue = sd(FINALVALUE, na.rm = TRUE),
    AverageValueWeightRatio = mean(value_weight_ratio, na.rm = TRUE),
    SDRatio = sd(value_weight_ratio, na.rm = TRUE),
    .groups = "drop"  # This removes the grouping structure after summarization
  )

print(SummaryStatsSpecies)
```


#Check Normality Assumptions
```{r}
# Test log transformed value-to-weight ratio for normality 
species_clean$ratio_log <- log(species_clean$value_weight_ratio)
ggplot(species_clean, aes(x = ratio_log))+
  geom_histogram()
shapiro.test(species_clean$ratio_log)
qqnorm(species_clean$ratio_log)
qqline(species_clean$ratio_log, col = "red")

# Test county data for normality 
ggplot(county_clean, aes(x = value_weight_ratio))+
  geom_histogram()
shapiro.test(county_clean$value_weight_ratio)
qqnorm(county_clean$value_weight_ratio)
qqline(county_clean$value_weight_ratio, col = "red")

# Apply the Shapiro-Wilk test on FINALWEIGHT to see if it is normal
#if p is less than 0.05, Not normal
shapiro_test_resultWeight <- shapiro.test(fish_all$FINALWEIGHT)
print(shapiro_test_resultWeight)

shapiro_test_resultValue <- shapiro.test(fish_all$FINALVALUE)
print(shapiro_test_resultValue)

#these are not normally distributed
```


```{r}
species_wilcox_results <- species_clean %>%
  filter(species_group != "Eels") %>%
  group_by(species_group) %>%
  summarise(p_value = wilcox.test(
    value_weight_ratio[Year == "2014-01-01"], 
    value_weight_ratio[Year == "2022-01-01"], 
    paired = TRUE,  # Indicates that the samples are paired
    alternative = "two.sided"  # Two-sided test
  )$p.value)

print(species_wilcox_results)

#shows us that other finfish, mackerals and tunas, groupers & seabass have all 
#changed in value from 2014 to 2022

#we use this test bc we don't assume normality
```

dfe3


```{r}
#are weight and value related? 

# so we are still interested in if weight and value are related 
species_correlation <- species_clean %>%
  filter(species_group != "Eels") %>%
  group_by(species_group, Year) %>%
  summarise(
    pearson_correlation = cor(FINALWEIGHT, FINALVALUE, method = "pearson", use = "complete.obs"),
    .groups = "drop"
  )

print(species_correlation)
# yes those two things are related
```


**Linear Model of Value to Weight**
```{r SPECIES data ratio linear model, message=FALSE, warning=FALSE, echo=FALSE}
#none normal data but maybe we can get away with it
species_ratio_lm<-lm(data=species_clean, log(FINALWEIGHT) ~ log(FINALVALUE))

summary(species_ratio_lm)
#significance!!

#sincer our data is not normal, maybe we can still use linear modeling if the 
#residuals of the linear model are normally distributed 

# Get residuals
res <- residuals(species_ratio_lm)
# Test for normality
shapiro.test(res)

#they are not, we cannot use linear modeling

bptest(species_ratio_lm)
#really cannot do linear modeling
```



**Here is a GLM Model**
```{r}
#generalized linear model does not need normallity
species_glm <- glm(FINALWEIGHT ~ FINALVALUE, data = species_clean, family = Gamma(link = "log"))
summary(species_glm)
#okay this model is signifciant 

# Calculate residuals
species_clean$residuals <- residuals(species_glm, type = "deviance")

# Plot residuals
ggplot(species_clean, aes(x = FINALVALUE, y = residuals)) +
  geom_point(alpha = 0.5) +  # Scatter plot of residuals
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals vs FINALVALUE", x = "FINALVALUE", y = "Residuals") +
  theme_minimal() +
  labs(title ="Residuals of the GLM Are Systematically Low")

#these residuals are not random... this model is not ideal? but GLM residuals don't need to be random for a valid model

# Generate predictions from the GLM
species_clean$predicted_values <- predict(species_glm, type = "response")

ggplot(species_clean, aes(x = FINALVALUE, y = FINALWEIGHT, color = FISH)) +
  geom_point(alpha = 0.5) +  # Scatter plot of actual data
  geom_line(aes(y = predicted_values), color = "blue", size = 1) +  # Predicted values from GLM
  labs(title = "Predicted vs Actual FINALWEIGHT",
       x = "FINALVALUE", y = "FINALWEIGHT") +
  theme_minimal() +
  theme(legend.position = "bottom")

#yeah, this model fit looks rough, way over predicts at higher ends like 
#residuals shows us
```

**Now Trying a GAM Model**
```{r}
# Fit a GAM model
species_gam <- gam(FINALWEIGHT ~ s(FINALVALUE), data = species_clean)

# Check the summary of the model
summary(species_gam)

# Generate predictions from the GAM model
species_clean$gam_predicted_values <- predict(species_gam, type = "response")

# Plot actual vs predicted using GAM
ggplot(species_clean, aes(x = FINALVALUE, y = FINALWEIGHT)) +
  geom_point(alpha = 0.5, color = "gray") +  # Scatter plot of actual data
  geom_line(aes(x = FINALVALUE, y = gam_predicted_values), color = "blue", size = 1) +  # GAM predicted values
  labs(title = "Predicted vs Actual FINALWEIGHT (GAM)", x = "FINALVALUE", y = "FINALWEIGHT") +
  theme_minimal()
#this model fit actually looks pretty good

# Compare AIC values
AIC(species_glm, species_gam)
# okay at this point we are probably overfitting the model ad this is a 
#result of uneccesary model completixty

# Calculate residuals
species_clean$residuals <- residuals(species_gam, type = "deviance")

# Plot residuals
ggplot(species_clean, aes(x = gam_predicted_values, y = residuals)) +
  geom_point(alpha = 0.5) +  # Scatter plot of residuals
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +  # Horizontal line at 0
  labs(title = "Residuals vs Predicted Values", x = "Predicted Values", y = "Residuals") +
  theme_minimal()

#wwell, these are random! that is good
```

**Now Trying Another GAM Model**
```{r}
# Fit a GAM with a categorical variable (species) 
species_gam2 <- gam(FINALWEIGHT ~ s(FINALVALUE) + species_group, data = species_clean)

# Check the summary of the model
summary(species_gam2)

# Predict the values using the fitted GAM model
species_clean$gam2_predicted_values <- predict(species_gam2, type = "response")

# Plot the actual vs predicted values
ggplot(species_clean, aes(x = FINALVALUE, y = FINALWEIGHT)) +
  geom_point(alpha = 0.5, color = "gray") +  # Scatter plot of actual data
  geom_line(aes(x = FINALVALUE, y = gam2_predicted_values), color = "blue", size = 1) +  # GAM predicted values
  labs(title = "Predicted vs Actual FINALWEIGHT (GAM)", x = "FINALVALUE", y = "FINALWEIGHT") +
  theme_minimal()

AIC(species_glm, species_gam, species_gam2)
```

**Now Trying Another GLM Model**
```{r}
species_glm2 <- glm(FINALWEIGHT ~ FINALVALUE * species_group, data = species_clean)
summary(species_glm2)

# Predict the values using the fitted GAM model
species_clean$glm2_predicted_values <- predict(species_glm2, type = "response")

# Plot the actual vs predicted values
ggplot(species_clean, aes(x = FINALVALUE, y = FINALWEIGHT)) +
  geom_point(alpha = 0.5, color = "gray") +  # Scatter plot of actual data
  geom_line(aes(x = FINALVALUE, y = glm2_predicted_values), color = "blue", size = 1) +  # GAM predicted values
  labs(title = "Predicted vs Actual FINALWEIGHT (GAM)", x = "FINALVALUE", y = "FINALWEIGHT") +
  theme_minimal()
```

```{r}
AIC(species_glm, species_gam, species_gam2, species_glm2)
#we are just overfitting now, the GLM model is the best option still
```

```{r}
kruskal_test <- kruskal.test(value_weight_ratio ~ species_group, data = species_clean)
print(kruskal_test)
# there are differences between at least one group and the median ratio value 
```

```{r}
pairwise.wilcox.test(
  x = species_clean$value_weight_ratio,
  g = species_clean$species_group,
  p.adjust.method = "BH"  # Benjamini-Hochberg, or use "bonferroni" or "holm"
)

```

```{r}
ggplot(species_clean, aes(x = species_group, y = value_weight_ratio)) +
  geom_boxplot(fill = "skyblue", alpha = 0.7, outlier.color = "red") +
  labs(
    title = "Value-to-Weight Ratio by Species Group",
    x = "Species Group",
    y = "Value-to-Weight Ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```




**MLR ratio across all years**
```{r SPECIES data multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
#log transform the variables to meet normality assumptions
species_multi_lm<-lm(data = species_clean, value_weight_ratio ~ Year + species_group)

fish_anova <- aov(value_weight_ratio ~ species_group, data = species_clean)
summary(fish_anova)
# There is a significant different in value across species groups.
fish_tukey <- TukeyHSD(fish_anova)
fish_tukey

summary(species_multi_lm)
plot(species_multi_lm)
summary(species_multi_lm)
```