---
title: "SchirmacherAyden_FinalProject"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
##Data Wrangling
```{r clear environment, include=FALSE, warning=FALSE, message=FALSE}
remove(list=ls())
```

```{r packages, warning=FALSE, message=FALSE, include=FALSE}
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("lubridate")
#install.packages("ggrepel")
#install.packages("here")
#install.packages("gifski")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("scales")
#install.packages("sf")
#install.packages("leaflet")
#install.packages("mapview")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggrepel)
library(here)
library(gifski)
library(gganimate)
library(gapminder)
library(scales)
library(sf)
library(leaflet)
library(mapview)
```

```{r data import, message=FALSE, warning=FALSE, include=FALSE}
getwd()

#catch and species information
fish_2022<-read.csv(here("Data", "Raw", "2022.csv"))
fish_2021<-read.csv(here("Data", "Raw", "2021.csv"))
fish_2020<-read.csv(here("Data", "Raw", "2020.csv"))
fish_2019<-read.csv(here("Data", "Raw", "2019.csv"))
fish_2018<-read.csv(here("Data", "Raw", "2018.csv"))
fish_2017<-read.csv(here("Data", "Raw", "2017.csv"))
fish_2016<-read.csv(here("Data", "Raw", "2016.csv"))
fish_2015<-read.csv(here("Data", "Raw", "2015.csv"))
fish_2014<-read.csv(here("Data", "Raw", "2014.csv"))

fish_all <- bind_rows(
  "2014" = fish_2014, 
  "2015" = fish_2015, 
  "2016" = fish_2016, 
  "2017" = fish_2017, 
  "2018" = fish_2018, 
  "2019" = fish_2019, 
  "2020" = fish_2020, 
  "2021" = fish_2021, 
  "2022" = fish_2022, 
  .id = "Year")


#spatial and county data
fish_10year<-read.csv(here("Data", "Raw", "stats_fishies.csv"))
fish_10year

```

```{r SPECIES data cleaning, message=FALSE, warning=FALSE, include=FALSE}

fish_all$FINALWEIGHT <- gsub("[,]", "", fish_all$FINALWEIGHT) 
#^^remove commas from data
fish_all$FINALVALUE <- gsub("[,]", "", fish_all$FINALVALUE)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("\\$", "", fish_all$FINALVALUE)
#^^remove dollar signs from data

fish_all<-fish_all%>%
  mutate(FINALWEIGHT=as.integer(FINALWEIGHT))%>% #make data into numerical
  mutate(FINALVALUE=as.integer(FINALVALUE))%>% #make data into numerical
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  #^^create date object
  filter(!is.na(FINALWEIGHT))%>% #filter for NA
  filter(!is.na(FINALVALUE)) #filter for NA

#Keeping only the fish observations that are present across all years
species_clean<-fish_all %>%
  select(Year, FISH, FINALWEIGHT, FINALVALUE)%>%
  group_by(FISH) %>%
  filter(n_distinct(Year) == length(unique(fish_all$Year))) %>%
  #^^keep only the observations across all years
  ungroup()
```

```{r COUNTY data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#group data by year and county
county_clean<-fish_10year%>%
  filter(!is.na(Pounds))%>% #filter for NA
  filter(!is.na(Value))%>% #filter for NA
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  group_by(County)%>%
  filter(n_distinct(Year) == length(unique(fish_10year$Year)))
  #^^keep only observations in all years
```


```{r SPECIES data group by species type, message=FALSE, warning=FALSE, include=FALSE}

species_clean <- species_clean %>%
  mutate(species_group = case_when(
    FISH %in% c("Blue Crabs, Hard", "Blue Crabs, Peeler", 
                "Blue Crabs, Soft", "Stone Crabs", ("Shrimp (Heads On)")) 
    ~ "Crustaceans",
    
    FISH %in% c("Clams, Hard (Meats)", "Clams, Hard (Numbers)", 
                "Oysters (Bushels)", "Oysters (Meats)", 
                "Scallop, Sea (Meats)", "Squid", 
                "Whelks/Conchs (Meats)", "Octopus") 
    ~ "Mollusks",

    FISH %in% c("Dogfish, Smooth", "Dogfish, Spiny", 
                "Sharks", "Skates") 
    ~ "Sharks & Rays",

    FISH %in% c("Eel, American") ~ "Eels",

    FISH %in% c("Flounder, Southern", "Flounder, Summer", 
                "Flounders, Other") 
    ~ "Flatfish",

    FISH %in% c("Croaker, Atlantic", "Drum, Black", "Drum, Red", 
                "Seatrout, Spotted", "Weakfish (Grey Trout)", 
                "Spot") 
    ~ "Drum & Croaker Family",

    FISH %in% c("Grouper, Gag", "Grouper, Red", "Grouper, Scamp", 
                "Grouper, Snowy", "Groupers, Other", "Sea Basses") 
    ~ "Groupers & Sea Basses",

    FISH %in% c("Snapper, Vermilion (Beeliner)", 
                "Snappers, Other", "Porgies") 
    ~ "Snappers & Porgies",

    FISH %in% c("Mackerel, Atlantic (Boston)", "Mackerel, King", 
                "Mackerel, Spanish", "Tuna, Bigeye", "Tuna, Bluefin",
                "Tuna, Yellowfin", "Tunas, Other", 
                "Tunny, Little (False Albacore)") ~ "Mackerels & Tunas",

    FISH %in% c("Amberjacks", 
                "Jacks (Crevalle, Rainbowrunner, Bluerunner)", "Pompano") 
    ~ "Jacks & Pompanos",

    FISH %in% c("Anglerfish (Monkfish Including Monklivers)", "Bluefish", "Bonito", 
                "Butterfish", "Carp", "Catfishes", "Cobia", "Cutlassfish, Atlantic", 
                "Dolphinfish", "Garfish", "Grunts", "Hakes", "Harvestfish (Starbutters)", 
                "Hogfish (Hog Snapper)", "Mullet, Sea (Kingfishes)", "Mullet, Striped", 
                "Perch, White", "Perch, Yellow", "Pigfish", "Pinfish", "Pufferfish", 
                "Scup", "Shad, American", "Shad, Gizzard", "Shad, Hickory", "Sheepshead", 
                "Skippers", "Spadefish", "Striped Bass", "Swordfish", "Tilefish, Blueline", 
                "Tilefish, Other", "Triggerfish", "Wahoo") ~ "Other Finfish",
    FISH %in% c("Unclassified Fish For Industrial/Bait", "Unclassified Fish for Food", 
                "Unclassified Shellfish") ~ "Unclassified",
    TRUE ~ "Other"
  ))

```

```{r SPECIES data new value to weight ratio, message=FALSE, warning=FALSE, include=FALSE}
species_clean<-species_clean%>%
  mutate(value_weight_ratio=FINALVALUE/FINALWEIGHT)
```


```{r SPECIES data create difference in means between 1972 and 2022, message=FALSE, warning=FALSE, include=FALSE}
species_year01<-species_clean%>%
  filter(Year=="2014-01-01")
species_year10<-species_clean%>%
  filter(Year=="2022-01-01")

species_diff<- 
  left_join(species_year10, species_year01, by = "FISH", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

species_clean <- left_join(species_clean, species_diff %>% 
                        select(FISH, diff_means), by = "FISH")

```

```{r COUNTY data create difference in means between 2014 and 2022, message=FALSE, warning=FALSE, include=FALSE}
county_clean<-county_clean%>%
  mutate(value_weight_ratio=Value/Pounds)

county_year01<-county_clean%>%
  filter(Year=="2014-01-01")
county_year10<-county_clean%>%
  filter(Year=="2022-01-01")

county_diff<- 
  left_join(county_year10, county_year01, by = "County", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

county_clean <- left_join(county_clean, county_diff 
                      %>% select(County, diff_means), by = "County")

```



##Exploratory Plots -- Catch and Species Data

*Average Weight-Value Ratio by Species Over Time*
```{r}
# Mean value ($) across species groups 
summary_table <- species_clean %>%
  group_by(Year, species_group) %>%
  summarize(Mean = mean(value_weight_ratio, na.rm = TRUE),
            Median = median(value_weight_ratio, na.rm = TRUE),
            SD = sd(value_weight_ratio, na.rm = TRUE))

# Print the summary table
print(summary_table)

# Visualization 
ggplot(summary_table, aes(x = Year, y = Mean, color = species_group)) +
  geom_line() +  # Line plot for the mean over time
  geom_point() +  # Add points to show individual data points
  labs(
    title = "Mean Weight-Value Ratio by Species Group over Time",
    x = "Time",
    y = "Mean Weight-Value Ratio"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 
 #+ facet_wrap(~species_group)
  ####!!! without the facet wrap the graph seems too busy, but up to you to decide!!
```

*Catch Weight-to-Value Data Categorized by Species*
```{r SPECIES data ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value from Commercial Fishing
    by Species Type in North Carolina, 2014-2022",
    x="Weight (lbs)", y="Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```

*Catch Weight-to-Value Data Categorized by Species with Log Transformations on X and Y Axis*
```{r SPECIES data log ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  scale_x_log10(labels=comma)+
  scale_y_log10(labels=comma)+
  geom_point(alpha = 0.40)+
  ggtitle("Logged Ratio of Fish Weight to Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  xlab("Log Scale Weight (lbs)")+
  ylab("Log Scale Value ($USD)")+
  theme_bw()

```

*ANIMATED Annual Catch Weight-to-Value Data Categorized by Species with Log Transformations on X and Y Axis and Labels on Outlying Values*
```{r SPECIES data animation over time, warning=FALSE, message=FALSE, echo=FALSE}
species_animate<-
  ggplot(species_clean, aes(x = FINALWEIGHT, y = FINALVALUE, 
         size=3,
         col = FISH)) +
    geom_point(alpha = 0.7, show.legend = FALSE) +
    scale_x_log10(labels=comma)+
    scale_y_continuous(labels=comma)+
    labs(title = 'Ratio of Value to Log Weight, Year: {closest_state}', 
        x = 'Log Scale Weight (Units)', 
        y = 'Value ($USD)') +
    transition_states(Year, transition_length = 1, state_length = 1,
                      wrap = TRUE)+
    ease_aes('cubic-in-out')+
    geom_text(data = species_clean%>%
      filter(FINALVALUE > 5000000),
      aes(label=FISH), 
      hjust = 1, vjust = 0, size = 3, color = "black")

animate(species_animate, fps = 30, duration = 15)
```

*Catch Weight-to-Value Data Categorized by County*
```{r COUNTY shapefile and joining dataset, warning=FALSE, message=FALSE, include=FALSE}
nc_county<-st_read(here("./Data/Processed/Spatial/cb_2018_us_county_20m.shp"))%>%
  filter(STATEFP == "37")

county_clean<-county_clean%>%
  mutate(NAME=County)

nc_county<-nc_county%>%
  inner_join(county_clean, by="NAME")

ggplot(nc_county)+
  geom_sf(aes(fill = value_weight_ratio))+
    labs(
    title= "Weight-Value Ratio of Commercial Seafood Landings by
    County of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```

*Catch Data County Spatial Plots Categorized by Value*
```{r COUNTY data spatial value plot, message=FALSE, warning=FALSE, echo=FALSE}
#ggplot(nc_county)+
#  geom_sf(aes(fill = Value))+
#    labs(
#    title= "Value of Commercial Seafood Landings by County
#    of Landing in North Carolina, 2014-2023")+
#  theme_bw()+
#  facet_wrap(~Year)
```

*Catch Data County Spatial Plots Categorized by Weight*
```{r COUNTY data spatial weight plot, message=FALSE, warning=FALSE, echo=FALSE}
#ggplot(nc_county)+
#  geom_sf(aes(fill = Pounds))+
#    labs(
#    title= "Weight of Commercial Seafood Landings by County
#    of Landing in North Carolina, 2014-2023")+
#  theme_bw()+
#  facet_wrap(~Year)
```

##Statistical Analysis -- Catch and Species Data

```{r}
#NEED SUMMARY STATISTICS
#NEED WILCOX RANKED TEST
```

**Two sample t-test for value-to-weight ratio 2014 vs. 2022**
```{r SPECIES data two sample t test between year 01 and year 10, message=FALSE, warning=FALSE, echo=FALSE}

species_t_test_results <- species_clean %>%
  filter(species_group !="Eels")%>%
  group_by(species_group) %>%
  summarise(p_value = t.test(
    value_weight_ratio[Year == "2014-01-01"], 
    value_weight_ratio[Year == "2022-01-01"], 
    alternative = "two.sided", 
    var.equal = FALSE
  )$p.value)

print(species_t_test_results)

#two sample t-test with the output showing the p-values... species groups with 
#significant p-values have a significant change between 2014 and 2022 in the
#value to weight ratio. 

#What do you mean? Just because the P-values change a lot does not make that significant...?

```


**Linear Model of Value to Weight**
```{r SPECIES data ratio linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_ratio_lm<-lm(data=fish_all, log(FINALWEIGHT) ~ log(FINALVALUE))

summary(species_ratio_lm)
#significance!!
```


**Linear Model Value to Weight Ratio across all years**
```{r SPECIES data value linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_yearratio_lm<-lm(data=species_clean, value_weight_ratio ~ Year) 
summary(species_yearratio_lm)


fish_anova <- aov(value_weight_ratio ~ species_group, data = species_clean)
summary(fish_anova)
# There is a significant different in value across species groups.
fish_tukey <- TukeyHSD(fish_anova)
fish_tukey
```

**MLR ratio across all years**
```{r SPECIES data multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
#log transform the variables to meet normality assumptions
species_multi_lm<-lm(data = species_clean, log(FINALVALUE) ~ log(FINALWEIGHT) + Year + species_group)

summary(species_multi_lm)
plot(species_multi_lm)
summary(species_multi_lm)
```