---
title: "SchirmacherAyden_FinalProject"
output:
  pdf_document: default
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
---
##Data Wrangling
```{r clear environment, include=FALSE, warning=FALSE, message=FALSE}
remove(list=ls())
```

```{r packages, warning=FALSE, message=FALSE, include=FALSE}
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("lubridate")
#install.packages("ggrepel")
#install.packages("here")
#install.packages("gifski")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("scales")
#install.packages("sf")
#install.packages("leaflet")
#install.packages("mapview")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggrepel)
library(here)
library(gifski)
library(gganimate)
library(gapminder)
library(scales)
library(sf)
library(leaflet)
library(mapview)
```

```{r data import, message=FALSE, warning=FALSE, include=FALSE}
getwd()

#catch and species information
fish_2022<-read.csv(here("Data", "Raw", "2022.csv"))
fish_2021<-read.csv(here("Data", "Raw", "2021.csv"))
fish_2020<-read.csv(here("Data", "Raw", "2020.csv"))
fish_2019<-read.csv(here("Data", "Raw", "2019.csv"))
fish_2018<-read.csv(here("Data", "Raw", "2018.csv"))
fish_2017<-read.csv(here("Data", "Raw", "2017.csv"))
fish_2016<-read.csv(here("Data", "Raw", "2016.csv"))
fish_2015<-read.csv(here("Data", "Raw", "2015.csv"))
fish_2014<-read.csv(here("Data", "Raw", "2014.csv"))

fish_all <- bind_rows(
  "2014" = fish_2014, 
  "2015" = fish_2015, 
  "2016" = fish_2016, 
  "2017" = fish_2017, 
  "2018" = fish_2018, 
  "2019" = fish_2019, 
  "2020" = fish_2020, 
  "2021" = fish_2021, 
  "2022" = fish_2022, 
  .id = "Year")


#spatial and county data
fish_10year<-read.csv(here("Data", "Raw", "stats_fishies.csv"))
fish_10year

```

```{r SPECIES data cleaning, message=FALSE, warning=FALSE, include=FALSE}

fish_all$FINALWEIGHT <- gsub("[,]", "", fish_all$FINALWEIGHT) 
#^^remove commas from data
fish_all$FINALVALUE <- gsub("[,]", "", fish_all$FINALVALUE)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("\\$", "", fish_all$FINALVALUE)
#^^remove dollar signs from data

fish_all<-fish_all%>%
  filter(!is.na(FINALWEIGHT))%>% #filter for NA
  filter(!is.na(FINALVALUE))%>% #filter for NA
  mutate(FINALWEIGHT=as.integer(FINALWEIGHT))%>% #make data into numerical
  mutate(FINALVALUE=as.integer(FINALVALUE))%>% #make data into numerical
 mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d")) 
#^^create date object

#Keeping only the fish observations that are present across all years
fish_all<-fish_all %>%
  select(Year, FISH, FINALWEIGHT, FINALVALUE)%>%
  group_by(FISH) %>%
  filter(n_distinct(Year) == length(unique(fish_all$Year))) %>%
  #^^keep only the observations across all years
  ungroup()
```

```{r COUNTY data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#group data by year and county
fish_clean<-fish_10year%>%
  filter(!is.na(Pounds))%>% #filter for NA
  filter(!is.na(Value))%>% #filter for NA
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  group_by(County)%>%
  filter(n_distinct(Year) == length(unique(fish_10year$Year)))
  #^^keep only observations in all years
fish_clean
```


```{r SPECIES data group by species type, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)

fish_all <- fish_all %>%
  mutate(species_group = case_when(
    FISH %in% c("Blue Crabs, Hard", "Blue Crabs, Peeler", 
                "Blue Crabs, Soft", "Stone Crabs", ("Shrimp (Heads On)")) 
    ~ "Crustaceans",
    
    FISH %in% c("Clams, Hard (Meats)", "Clams, Hard (Numbers)", 
                "Oysters (Bushels)", "Oysters (Meats)", 
                "Scallop, Sea (Meats)", "Squid", 
                "Whelks/Conchs (Meats)", "Octopus") 
    ~ "Mollusks",

    FISH %in% c("Dogfish, Smooth", "Dogfish, Spiny", 
                "Sharks", "Skates") 
    ~ "Sharks & Rays",

    FISH %in% c("Eel, American") ~ "Eels",

    FISH %in% c("Flounder, Southern", "Flounder, Summer", 
                "Flounders, Other") 
    ~ "Flatfish",

    FISH %in% c("Croaker, Atlantic", "Drum, Black", "Drum, Red", 
                "Seatrout, Spotted", "Weakfish (Grey Trout)", 
                "Spot") 
    ~ "Drum & Croaker Family",

    FISH %in% c("Grouper, Gag", "Grouper, Red", "Grouper, Scamp", 
                "Grouper, Snowy", "Groupers, Other", "Sea Basses") 
    ~ "Groupers & Sea Basses",

    FISH %in% c("Snapper, Vermilion (Beeliner)", 
                "Snappers, Other", "Porgies") 
    ~ "Snappers & Porgies",

    FISH %in% c("Mackerel, Atlantic (Boston)", "Mackerel, King", 
                "Mackerel, Spanish", "Tuna, Bigeye", "Tuna, Bluefin",
                "Tuna, Yellowfin", "Tunas, Other", 
                "Tunny, Little (False Albacore)") ~ "Mackerels & Tunas",

    FISH %in% c("Amberjacks", 
                "Jacks (Crevalle, Rainbowrunner, Bluerunner)", "Pompano") 
    ~ "Jacks & Pompanos",

    FISH %in% c("Anglerfish (Monkfish Including Monklivers)", "Bluefish", "Bonito", 
                "Butterfish", "Carp", "Catfishes", "Cobia", "Cutlassfish, Atlantic", 
                "Dolphinfish", "Garfish", "Grunts", "Hakes", "Harvestfish (Starbutters)", 
                "Hogfish (Hog Snapper)", "Mullet, Sea (Kingfishes)", "Mullet, Striped", 
                "Perch, White", "Perch, Yellow", "Pigfish", "Pinfish", "Pufferfish", 
                "Scup", "Shad, American", "Shad, Gizzard", "Shad, Hickory", "Sheepshead", 
                "Skippers", "Spadefish", "Striped Bass", "Swordfish", "Tilefish, Blueline", 
                "Tilefish, Other", "Triggerfish", "Wahoo") ~ "Other Finfish",
    FISH %in% c("Unclassified Fish For Industrial/Bait", "Unclassified Fish for Food", 
                "Unclassified Shellfish") ~ "Unclassified",
    TRUE ~ "Other"
  ))

```

```{r SPECIES data new value to weight ratio, message=FALSE, warning=FALSE, include=FALSE}
fish_all<-fish_all%>%
  mutate(value_weight_ratio=FINALVALUE/FINALWEIGHT)
```


```{r SPECIES data create difference in means between 1972 and 2022, message=FALSE, warning=FALSE, include=FALSE}
species_year01<-fish_all%>%
  filter(Year=="2014-01-01")
species_year10<-fish_all%>%
  filter(Year=="2022-01-01")

species_diff<- 
  left_join(species_year10, species_year01, by = "FISH", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

fish_all <- left_join(fish_all, species_diff %>% 
                        select(FISH, diff_means), by = "FISH")

```

```{r COUNTY data create difference in means between 2014 and 2022, message=FALSE, warning=FALSE, include=FALSE}
fish_clean<-fish_clean%>%
  mutate(value_weight_ratio=Value/Pounds)

county_year01<-fish_clean%>%
  filter(Year=="2014-01-01")
county_year10<-fish_clean%>%
  filter(Year=="2022-01-01")

county_diff<- 
  left_join(county_year10, county_year01, by = "County", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

fish_clean <- left_join(fish_clean, county_diff 
                      %>% select(County, diff_means), by = "County")


```



##Exploratory Plots -- Catch and Species Data

```{r SPECIES data log plot value by species group, warning=FALSE, message=FALSE, echo=FALSE, fig.height=12, fig.width=8}
ggplot(fish_all, aes(x = Year, y = FINALVALUE, col = species_group))+
  geom_point()+
  scale_y_log10()+
  geom_smooth(method = lm, color="red", linewidth=0.45)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))+
  labs(x="Year", y="Value ($USD)", 
  title = "Final Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  facet_wrap(~ species_group, ncol=2)
```

```{r SPECIES data ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value from Commercial Fishing
    by Species Type in North Carolina, 2014-2022",
    x="Weight (lbs)", y="Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```

```{r SPECIES data log ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  scale_x_log10(labels=comma)+
  scale_y_log10(labels=comma)+
  geom_point(alpha = 0.40)+
  ggtitle("Logged Ratio of Fish Weight to Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  xlab("Log Scale Weight (lbs)")+
  ylab("Log Scale Value ($USD)")+
  theme_bw()

```

```{r SPECIES data animation over time, warning=FALSE, message=FALSE, echo=FALSE}
#this animation plot can't be knitted to a PDF file
#fish_animate<-
  #ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, 
         #              size=3,
         #              col = FISH)) +
    #geom_point(alpha = 0.7, show.legend = FALSE) +
    #scale_x_log10(labels=comma)+
   # scale_y_continuous(labels=comma)+
   # labs(title = 'Ratio of Value to Log Weight, Year: {closest_state}', 
    #   x = 'Log Scale Weight (Units)', 
    #   y = 'Value ($USD)') +
    #transition_states(Year, transition_length = 1, state_length = 1,
   #                   wrap = TRUE)+
    #ease_aes('cubic-in-out')+
  #geom_text(data = fish_all%>%
  #  filter(FINALVALUE > 5000000),
  #  aes(label=FISH), 
  #  hjust = 1, vjust = 0, size = 3, color = "black")

#animate(fish_animate, fps = 30, duration = 15)
```


<<<<<<< HEAD
**Basic Stats**

```{r}
# Mean value ($) across species groups 
summary_table <- fish_all %>%
  group_by(Year, species_group) %>%
  summarize(Mean = mean(FINALVALUE, na.rm = TRUE),
            Median = median(FINALVALUE, na.rm = TRUE),
            SD = sd(FINALVALUE, na.rm = TRUE))

# Print the summary table
print(summary_table)

# Visualization 
ggplot(summary_table, aes(x = Year, y = Mean, color = species_group)) +
  geom_line() +  # Line plot for the mean over time
  geom_point() +  # Add points to show individual data points
  labs(
    title = "Mean Value over Time by Species Group",
    x = "Time",
    y = "Mean Value ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 

# Removing crustaceans 
summary_data <- summary_table %>% 
  filter(species_group != "Crustaceans")

ggplot(summary_data, aes(x = Year, y = Mean, color = species_group)) +
  geom_line() +  # Line plot for the mean over time
  geom_point() +  # Add points to show individual data points
  labs(
    title = "Mean Value over Time by Species Group",
    x = "Time",
    y = "Mean Value ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "right") 
```

```{r}
# Check for normality 
ggplot(fish_all, aes(sample = FINALVALUE)) +
  geom_qq() + 
  geom_qq_line() +
  facet_wrap(~species_group)

```


**Two sample t-test for value-to-weight ratio 1972 vs. 2022**
```{r two sample t test between year 01 and year 50, message=FALSE, warning=FALSE, echo=FALSE}
=======
##Exploratory Plots -- Spatial and County Data
>>>>>>> 60d4990448a059ac94f542bee6392c62912efbf6

```{r COUNTY data ratio plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(fish_clean, aes(x = Pounds, y = Value, col=County))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value from Commercial Fishing
    by County in North Carolina, 2014-2022",
    x="Weight (lbs)", y="Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```

```{r COUNTY log data ratio plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(fish_clean, aes(x = log(Pounds), y = log(Value), col=County))+
  geom_point(alpha = 0.40)+
  labs(
    title="Logged Ratio of Fish Weight to Value from Commercial Fishing
    by County in North Carolina, 2014-2022",
    x="Log Scale Weight (lbs)", y="Log Scale Value ($USD)")+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```




```{r COUNTY shapefile and joining dataset, warning=FALSE, message=FALSE, include=FALSE}
nc_county<-st_read(here("./Data/Processed/Spatial/cb_2018_us_county_20m.shp"))%>%
  filter(STATEFP == "37")

fish_clean<-fish_clean%>%
  mutate(NAME=County)

nc_county<-nc_county%>%
  inner_join(fish_clean, by="NAME")

```


```{r COUNTY data spatial value plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = Value))+
    labs(
    title= "Value of Commercial Seafood Landings by County
    of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```

```{r COUNTY data spatial weight plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = Pounds))+
    labs(
    title= "Weight of Commercial Seafood Landings by County
    of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```


```{r COUNTY weight-value ratio spatial plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = value_weight_ratio))+
    labs(
    title= "Weight-Value Ratio of Commercial Seafood Landings by
    County of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```


```{r COUNTY weight-value ratio spatial plot 2014 vs. 2023, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = diff_means))+
    labs(
    title= "Difference in Weight-Value Ratio of Commercial 
    Seafood Landings Between 2014 and 2023 by
    County of Landing in North Carolina")+
  theme_bw()
```



##Statistical Analysis -- Catch and Species Data

**Two sample t-test for value-to-weight ratio 2014 vs. 2022**
```{r SPECIES data two sample t test between year 01 and year 10, message=FALSE, warning=FALSE, echo=FALSE}

species_t_test_results <- fish_all %>%
  filter(species_group !="Eels")%>%
  group_by(species_group) %>%
  summarise(p_value = t.test(
    value_weight_ratio[Year == "2014-01-01"], 
    value_weight_ratio[Year == "2022-01-01"], 
    alternative = "two.sided", 
    var.equal = FALSE
  )$p.value)

print(species_t_test_results)

#two sample t-test with the output showing the p-values... species groups with 
#significant p-values have a significant change between 2014 and 2022 in the
#value to weight ratio. 

```

**Linear Model Value across all years**
```{r SPECIES data value linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_value_lm<-lm(data=fish_all, FINALVALUE ~ Year) 

<<<<<<< HEAD
summary(fish_anova)

fish_anova2 <- aov(FINALVALUE ~ species_group, data = fish_all)
summary(fish_anova2)
# There is a significant different in value across species groups.
fish_tukey <- TukeyHSD(fish_anova2)
fish_tukey

# There is a significant difference in value between the following species groups:
# Bivalves significantly different from Drums & Croakers, Other common finfish,
# Perches, Shads, and Unclassified.
# Crustaceans significantly different than all other species groups.
# Flatfish significantly different from Drums & Croakers, Mackerels, Mullets,
# other, other common finfish, Perches, Shads, and unclassified (all except
# bivalves). 
# Other common finfish significantly different from others.

=======
summary(species_value_lm)
>>>>>>> 60d4990448a059ac94f542bee6392c62912efbf6
```

**Linear Model Weight across all years**
```{r SPECIES data weight linear model, message=FALSE, warning=FALSE, echo=FALSE}
fish_weight_lm<-lm(data=fish_all, FINALWEIGHT ~ Year) 

<<<<<<< HEAD
summary(fish_lm)
plot(fish_lm)
=======
summary(fish_weight_lm)
  #significance!!
```

**Linear Model RATIO across all years**
```{r SPECIES data ratio linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_ratio_lm<-lm(data=fish_all, log(FINALWEIGHT) ~ log(FINALVALUE))

summary(species_ratio_lm)
#significance!!
>>>>>>> 60d4990448a059ac94f542bee6392c62912efbf6
```

**MLR ratio across all years**
```{r SPECIES data multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
#log transform the variables to meet normality assumptions
species_multi_lm<-lm(data = fish_all, log(FINALVALUE) ~ log(FINALWEIGHT) + Year + species_group)

<<<<<<< HEAD
summary(fish_multi_lm)
plot(fish_multi_lm)
=======
summary(species_multi_lm)
```

##Statistical Analysis -- County and Spatial Data

**Two sample t-test for value-to-weight ratio 2014 vs. 2022**
```{r COUNTY data two sample t test between year 01 and year 10, message=FALSE, warning=FALSE, echo=FALSE}

#I CANNOT GET THIS TO WORK

#this does not work becase you only have 1 observation per group I think, cannot use a t-test then

#county_t_test_results <- fish_clean %>%
  #group_by(County) %>%
  #summarise(p_value = t.test(
  #  value_weight_ratio[Year == "2014-01-01"], 
  #  value_weight_ratio[Year == "2022-01-01"], 
  #  alternative = "two.sided", 
  #  var.equal = FALSE
 # )$p.value)

#print(t_test_results)

#two sample t-test with the output showing the p-values... counties with 
#significant p-values have a significant change between 2014 and 2022 in the
#value to weight ratio. 



fish_clean %>%
  filter(Year %in% as.Date(c("2014-01-01", "2022-01-01"))) %>%
  group_by(County, Year) %>%
  summarise(n = sum(!is.na(value_weight_ratio)), .groups = "drop")




```

**Linear Model Value across all years**
```{r COUNTY value linear model, message=FALSE, warning=FALSE, echo=FALSE}
county_value_lm<-lm(data=fish_clean, Value ~ Year) 

summary(county_value_lm)
```

**Linear Model Weight across all years**
```{r COUNTY weight linear model, message=FALSE, warning=FALSE, echo=FALSE}
county_weight_lm<-lm(data=fish_clean, Pounds ~ Year) 

summary(county_weight_lm)
  #significance!!
```

**Linear Model RATIO across all years**
```{r COUNTY ratio linear model, message=FALSE, warning=FALSE, echo=FALSE}
county_ratio_lm<-lm(data=fish_clean, log(Pounds) ~ log(Value)) 

summary(county_ratio_lm)
#significance!!
```

**MLR ratio across all years**
```{r COUNTY multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
county_multi_lm<-lm(data = fish_clean, Value ~ Pounds + Year + County)

#remembering that for this the intercept is Beaufort!!

summary(county_multi_lm)
>>>>>>> 60d4990448a059ac94f542bee6392c62912efbf6
```


