---
title: "SchirmacherAyden_FinalProject"
output:
  html_document:
    df_print: paged
date: "`r Sys.Date()`"
---
##Data Wrangling
```{r clear environment, include=FALSE, warning=FALSE, message=FALSE}
remove(list=ls())
```

```{r packages, warning=FALSE, message=FALSE, include=FALSE}
#install.packages("dplyr")
#install.packages("tidyverse")
#install.packages("ggplot2")
#install.packages("lubridate")
#install.packages("ggrepel")
#install.packages("here")
#install.packages("gifski")
#install.packages("gganimate")
#install.packages("gapminder")
#install.packages("scales")
#install.packages("sf")
#install.packages("leaflet")
#install.packages("mapview")
library(dplyr)
library(tidyverse)
library(ggplot2)
library(lubridate)
library(ggrepel)
library(here)
library(gifski)
library(gganimate)
library(gapminder)
library(scales)
library(sf)
library(leaflet)
library(mapview)
```

```{r data import, message=FALSE, warning=FALSE, include=FALSE}
getwd()

#catch and species information
fish_2022<-read.csv(here("Data", "Raw", "2022.csv"))
fish_2021<-read.csv(here("Data", "Raw", "2021.csv"))
fish_2020<-read.csv(here("Data", "Raw", "2020.csv"))
fish_2019<-read.csv(here("Data", "Raw", "2019.csv"))
fish_2018<-read.csv(here("Data", "Raw", "2018.csv"))
fish_2017<-read.csv(here("Data", "Raw", "2017.csv"))
fish_2016<-read.csv(here("Data", "Raw", "2016.csv"))
fish_2015<-read.csv(here("Data", "Raw", "2015.csv"))
fish_2014<-read.csv(here("Data", "Raw", "2014.csv"))

fish_all <- bind_rows(
  "2014" = fish_2014, 
  "2015" = fish_2015, 
  "2016" = fish_2016, 
  "2017" = fish_2017, 
  "2018" = fish_2018, 
  "2019" = fish_2019, 
  "2020" = fish_2020, 
  "2021" = fish_2021, 
  "2022" = fish_2022, 
  .id = "Year")


#spatial and county data
fish_10year<-read.csv(here("Data", "Raw", "stats_fishies.csv"))
fish_10year

```

```{r SPECIES data cleaning, message=FALSE, warning=FALSE, include=FALSE}

fish_all$FINALWEIGHT <- gsub("[,]", "", fish_all$FINALWEIGHT) 
#^^remove commas from data
fish_all$FINALVALUE <- gsub("[,]", "", fish_all$FINALVALUE)
#^^remove commas from data
fish_all$FINALVALUE <- gsub("\\$", "", fish_all$FINALVALUE)
#^^remove dollar signs from data

fish_all<-fish_all%>%
  filter(!is.na(FINALWEIGHT))%>% #filter for NA
  filter(!is.na(FINALVALUE))%>% #filter for NA
  mutate(FINALWEIGHT=as.integer(FINALWEIGHT))%>% #make data into numerical
  mutate(FINALVALUE=as.integer(FINALVALUE))%>% #make data into numerical
 mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d")) 
#^^create date object

#Keeping only the fish observations that are present across all years
fish_all<-fish_all %>%
  select(Year, FISH, FINALWEIGHT, FINALVALUE)%>%
  group_by(FISH) %>%
  filter(n_distinct(Year) == length(unique(fish_all$Year))) %>%
  #^^keep only the observations across all years
  ungroup()
```

```{r COUNTY data cleaning, message=FALSE, warning=FALSE, include=FALSE}
#group data by year and county
fish_clean<-fish_10year%>%
  filter(!is.na(Pounds))%>% #filter for NA
  filter(!is.na(Value))%>% #filter for NA
  mutate(Year = as.Date(paste0(Year, "-01-01"), format="%Y-%m-%d"))%>%
  group_by(County)%>%
  filter(n_distinct(Year) == length(unique(fish_10year$Year)))
  #^^keep only observations in all years
fish_clean
```


```{r SPECIES data group by species type, message=FALSE, warning=FALSE, include=FALSE}
library(dplyr)

fish_all <- fish_all %>%
  mutate(species_group = case_when(
    FISH %in% c("Blue Crabs, Hard", "Blue Crabs, Peeler", 
                "Blue Crabs, Soft", "Stone Crabs", ("Shrimp (Heads On)")) 
    ~ "Crustaceans",
    
    FISH %in% c("Clams, Hard (Meats)", "Clams, Hard (Numbers)", 
                "Oysters (Bushels)", "Oysters (Meats)", 
                "Scallop, Sea (Meats)", "Squid", 
                "Whelks/Conchs (Meats)", "Octopus") 
    ~ "Mollusks",

    FISH %in% c("Dogfish, Smooth", "Dogfish, Spiny", 
                "Sharks", "Skates") 
    ~ "Sharks & Rays",

    FISH %in% c("Eel, American") ~ "Eels",

    FISH %in% c("Flounder, Southern", "Flounder, Summer", 
                "Flounders, Other") 
    ~ "Flatfish",

    FISH %in% c("Croaker, Atlantic", "Drum, Black", "Drum, Red", 
                "Seatrout, Spotted", "Weakfish (Grey Trout)", 
                "Spot") 
    ~ "Drum & Croaker Family",

    FISH %in% c("Grouper, Gag", "Grouper, Red", "Grouper, Scamp", 
                "Grouper, Snowy", "Groupers, Other", "Sea Basses") 
    ~ "Groupers & Sea Basses",

    FISH %in% c("Snapper, Vermilion (Beeliner)", 
                "Snappers, Other", "Porgies") 
    ~ "Snappers & Porgies",

    FISH %in% c("Mackerel, Atlantic (Boston)", "Mackerel, King", 
                "Mackerel, Spanish", "Tuna, Bigeye", "Tuna, Bluefin",
                "Tuna, Yellowfin", "Tunas, Other", 
                "Tunny, Little (False Albacore)") ~ "Mackerels & Tunas",

    FISH %in% c("Amberjacks", 
                "Jacks (Crevalle, Rainbowrunner, Bluerunner)", "Pompano") 
    ~ "Jacks & Pompanos",

    FISH %in% c("Anglerfish (Monkfish Including Monklivers)", "Bluefish", "Bonito", 
                "Butterfish", "Carp", "Catfishes", "Cobia", "Cutlassfish, Atlantic", 
                "Dolphinfish", "Garfish", "Grunts", "Hakes", "Harvestfish (Starbutters)", 
                "Hogfish (Hog Snapper)", "Mullet, Sea (Kingfishes)", "Mullet, Striped", 
                "Perch, White", "Perch, Yellow", "Pigfish", "Pinfish", "Pufferfish", 
                "Scup", "Shad, American", "Shad, Gizzard", "Shad, Hickory", "Sheepshead", 
                "Skippers", "Spadefish", "Striped Bass", "Swordfish", "Tilefish, Blueline", 
                "Tilefish, Other", "Triggerfish", "Wahoo") ~ "Other Finfish",
    FISH %in% c("Unclassified Fish For Industrial/Bait", "Unclassified Fish for Food", 
                "Unclassified Shellfish") ~ "Unclassified",
    TRUE ~ "Other"
  ))

```

```{r SPECIES data new value to weight ratio, message=FALSE, warning=FALSE, include=FALSE}
fish_all<-fish_all%>%
  mutate(value_weight_ratio=FINALVALUE/FINALWEIGHT)
```

```{r SPECIES data create difference in means between 1972 and 2022, message=FALSE, warning=FALSE, include=FALSE}
species_year01<-fish_all%>%
  filter(Year=="2014-01-01")
species_year10<-fish_all%>%
  filter(Year=="2022-01-01")

species_diff<- 
  left_join(species_year10, species_year01, by = "FISH", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

fish_all <- left_join(fish_all, species_diff %>% 
                        select(FISH, diff_means), by = "FISH")

```

```{r COUNTY data create difference in means between 2014 and 2022, message=FALSE, warning=FALSE, include=FALSE}
fish_clean<-fish_clean%>%
  mutate(value_weight_ratio=Value/Pounds)

county_year01<-fish_clean%>%
  filter(Year=="2014-01-01")
county_year10<-fish_clean%>%
  filter(Year=="2022-01-01")

county_diff<- 
  left_join(county_year10, county_year01, by = "County", suffix = c("_2022", "_2014")) %>%
  mutate(diff_means = value_weight_ratio_2022 - value_weight_ratio_2014)

fish_clean <- left_join(fish_clean, county_diff 
                      %>% select(County, diff_means), by = "County")


```



##Exploratory Plots -- Catch and Species Data
```{r SPECIES data value by year plot, message=FALSE, warning=TRUE, echo=FALSE}
ggplot(fish_all, aes(x = Year, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Final Value from Commercial Fishing by Species Type
    in North Carolina, 2014-2022",
    x="Year", y="Value ($USD)")+
  theme_bw()
```

```{r SPECIES data weight by year plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = Year, y = FINALWEIGHT, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Final Weight from Commercial Fishing by Species Type
    in North Carolina, 2014-2022",
    x="Year", y="Weight (lbs)")+
  theme_bw()

```

```{r SPECIES data log plot value by species group, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = Year, y = FINALVALUE, col = species_group))+
  geom_point()+
  scale_y_log10()+
  geom_smooth(method = lm, color="red", linewidth=0.45)+
  labs(x="Year", y="Value ($USD)", 
  title = "Final Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  facet_wrap(~ species_group)
```

```{r SPECIES data ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  geom_point(alpha = 0.40)+
  labs(
    title="Ratio of Fish Weight to Value from Commercial Fishing
    by Species Type in North Carolina, 2014-2022",
    x="Weight (lbs)", y="Value ($USD)")+
  geom_text_repel(data = fish_all%>%
                  filter(abs(FINALVALUE) > 23000000),
                  aes(label = paste(FISH, format(Year, "%Y"))),
                  color = 'red', size=3)+
    scale_y_continuous(labels=comma)+
  scale_x_continuous(labels=comma)+
  theme_bw()
```

```{r SPECIES data log ratio plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, col=species_group))+
  scale_x_log10(labels=comma)+
  scale_y_log10(labels=comma)+
  geom_point(alpha = 0.40)+
  ggtitle("Logged Ratio of Fish Weight to Value from Commercial Fishing
          by Species Type in North Carolina, 2014-2022")+
  xlab("Log Scale Weight (lbs)")+
  ylab("Value ($USD)")+
  theme_bw()

```

```{r SPECIES data animation over time, warning=FALSE, message=FALSE, echo=FALSE}
#this animation plot can't be knitted to a PDF file
#fish_animate<-
  #ggplot(fish_all, aes(x = FINALWEIGHT, y = FINALVALUE, 
         #              size=3,
         #              col = FISH)) +
    #geom_point(alpha = 0.7, show.legend = FALSE) +
    #scale_x_log10(labels=comma)+
   # scale_y_continuous(labels=comma)+
   # labs(title = 'Ratio of Value to Log Weight, Year: {closest_state}', 
    #   x = 'Log Scale Weight (Units)', 
    #   y = 'Value ($USD)') +
    #transition_states(Year, transition_length = 1, state_length = 1,
   #                   wrap = TRUE)+
    #ease_aes('cubic-in-out')+
  #geom_text(data = fish_all%>%
  #  filter(FINALVALUE > 5000000),
  #  aes(label=FISH), 
  #  hjust = 1, vjust = 0, size = 3, color = "black")

#animate(fish_animate, fps = 30, duration = 15)
```


##Exploratory Plots -- Spatial and County Data
```{r COUNTY data spatial value plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = Value))+
    labs(
    title= "Value of Commercial Seafood Landings by County
    of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```

```{r COUNTY data spatial weight plot, message=FALSE, warning=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = Pounds))+
    labs(
    title= "Weight of Commercial Seafood Landings by County
    of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```

```{r COUNTY shapefile and joining dataset, warning=FALSE, message=FALSE, include=FALSE}
nc_county<-st_read(here("./Data/Processed/Spatial/cb_2018_us_county_20m.shp"))%>%
  filter(STATEFP == "37")

fish_clean<-fish_clean%>%
  mutate(NAME=County)

nc_county<-nc_county%>%
  inner_join(fish_clean, by="NAME")

```


```{r COUNTY weight-value ratio spatial plot, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = value_weight_ratio))+
    labs(
    title= "Weight-Value Ratio of Commercial Seafood Landings by
    County of Landing in North Carolina, 2014-2023")+
  theme_bw()+
  facet_wrap(~Year)
```


```{r COUNTY weight-value ratio spatial plot 2014 vs. 2023, warning=FALSE, message=FALSE, echo=FALSE}
ggplot(nc_county)+
  geom_sf(aes(fill = diff_means))+
    labs(
    title= "Difference in Weight-Value Ratio of Commercial 
    Seafood Landings Between 2014 and 2023 by
    County of Landing in North Carolina")+
  theme_bw()
```

```{r COUNTY value by summed totals, warning=FALSE, message=FALSE, echo=FALSE}

spatial_value_plot<-
  ggplot(nc_county)+
  geom_sf(aes(fill = difference_value))+
    labs(
    title="Difference in Value of Commercial Seafood Landings Between 
    2014 and 2023 by County of Landing in North Carolina")+
  theme_bw()

spatial_value_plot
```



##Statistical Analysis -- Catch and Species Data

**Two sample t-test for value-to-weight ratio 2014 vs. 2022**
```{r SPECIES data two sample t test between year 01 and year 10, message=FALSE, warning=FALSE, echo=FALSE}

species_t_test_results <- fish_all %>%
  filter(species_group !="Eels")%>%
  group_by(species_group) %>%
  summarise(p_value = t.test(
    value_weight_ratio[Year == "2014-01-01"], 
    value_weight_ratio[Year == "2022-01-01"], 
    alternative = "two.sided", 
    var.equal = FALSE
  )$p.value)

print(species_t_test_results)

#two sample t-test with the output showing the p-values... species groups with 
#significant p-values have a significant change between 2014 and 2022 in the
#value to weight ratio. 

```

**Linear Model Value across all years**
```{r SPECIES data linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_value_lm<-lm(data=fish_all, FINALVALUE ~ Year) 

summary(species_value_lm)
```

**Linear Model Weight across all years**
```{r SPECIES data linear model, message=FALSE, warning=FALSE, echo=FALSE}
fish_weight_lm<-lm(data=fish_all, FINALWEIGHT ~ Year) 

summary(fish_weight_lm)
  #significance!!
```

**Linear Model RATIO across all years**
```{r SPECIES data linear model, message=FALSE, warning=FALSE, echo=FALSE}
species_ratio_lm<-lm(data=fish_all, FINALWEIGHT ~ FINALVALUE) 

summary(species_ratio_lm)
#significance!!
```

**MLR ratio across all years**
```{r SPECIES data multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
species_multi_lm<-lm(data = fish_all, FINALVALUE ~ FINALWEIGHT + Year + species_group)

summary(species_multi_lm)
```

##Statistical Analysis -- County and Spatial Data

**Two sample t-test for value-to-weight ratio 2014 vs. 2022**
```{r COUNTY data two sample t test between year 01 and year 10, message=FALSE, warning=FALSE, echo=FALSE}

county_t_test_results <- fish_clean %>%
  group_by(County) %>%
  summarise(p_value = t.test(
    value_weight_ratio[Year == "2014-01-01"], 
    value_weight_ratio[Year == "2022-01-01"], 
    alternative = "two.sided", 
    var.equal = FALSE
  )$p.value)

print(t_test_results)

#two sample t-test with the output showing the p-values... species groups with 
#significant p-values have a significant change between 2014 and 2022 in the
#value to weight ratio. 

```

**Linear Model Value across all years**
```{r COUNTY linear model, message=FALSE, warning=FALSE, echo=FALSE}
fish_value_lm<-lm(data=fish_all, FINALVALUE ~ Year) 

summary(fish_value_lm)
```

**Linear Model Weight across all years**
```{r COUNTY linear model, message=FALSE, warning=FALSE, echo=FALSE}
fish_weight_lm<-lm(data=fish_all, FINALWEIGHT ~ Year) 

summary(fish_weight_lm)
  #significance!!
```

**Linear Model RATIO across all years**
```{r COUNTY linear model, message=FALSE, warning=FALSE, echo=FALSE}
fish_ratio_lm<-lm(data=fish_all, FINALWEIGHT ~ FINALVALUE) 

summary(fish_ratio_lm)
#significance!!
```

**MLR ratio across all years**
```{r COUNTY multiple linear regression model, message=FALSE, warning=FALSE, echo=FALSE}
fish_multi_lm<-lm(data = fish_all, FINALVALUE ~ FINALWEIGHT + Year + species_group)

summary(fish_multi_lm)
```


